\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2021-08-08 manuscript.Rnw
Minor text edits to manuscript

There are a number of multi-state modeling \proglang{R} packages that can both fit models and make predictions. \pkg{hesim} leverages some of these for parameters estimation: \pkg{flexsurv} is used to fit parametric and flexible parametric survival and multi-state models when exact event times are available while \pkg{msm} is used to fit multi-state models when only panel data is available. A key difference is that \pkg{hesim} is designed specifically for health economic modeling whereas the other packages are more general purpose statistical packages.

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models \citep{de2011mstate}. It can predict state probabilities and simulate multi-state trajectories (to facilitate prediction for clock-reset models) from cumulative hazard functions evaluated at specific times. \citet{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health economic modeling since it was not designed for it. \code{mssample()}---the core function for simulating multi-state models---does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is also less computationally efficient than \pkg{hesim} for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions when efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance, we ran (see ``Benchmarks'' on the package website) 1,000 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. At the time of writing, the model completed in 2.8 seconds in \pkg{hesim} and 3.1 hours in \pkg{mstate}, showing that computational considerations are not just an academic concern.

\pkg{heemod} is the only other general purpose \proglang{R} package for cost-effectiveness modeling. As noted in the introduction, there are some important differences between \pkg{hesim} and \pkg{heemod} related to the supported model types and integration of statistical and economic modeling. In addition, as with \pkg{mstate}, simulations are not vectorized, which could result in significantly reduced performance. To test this, we simulated a time-inhomogeneous total hip replacement example from \citet{briggs2006decision}. A cohort model in \pkg{heemod} completed in approximately 1.9 minutes while the same cohort model took approximately 1 second to run in \pkg{hesim} and an equivalent individual-level simulation in \pkg{hesim} ran in 7.4 seconds.

There are of course features in other \proglang{R} packages that are not in the current version of \pkg{hesim}. For example, \pkg{heemod} has an explicit function, \code{calibrate_model()}, for calibrating model parameters that are difficult to measure. Furthermore, tunnel states can be automatically created (rather than needing to be created manually) with \pkg{heemod} using the \code{state_time} variable so that costs or transition probabilities can very as a function of time in a health state. \pkg{hesim} is also limited in its support for value of information analyses. In particular, neither the expected value of partial perfect information (EVPPI) or the expected value of sample information (EVSI) can be computed with \pkg{hesim}, whereas the EVPPI can be computed with \pkg{BCEA}, \pkg{dampack}, and \pkg{SAVI} and EVSI can be computed with \pkg{EVSI} and \pkg{dampack}. In this respect, these packages are complementary and simulation output from \pkg{hesim} could be used as inputs to their more advanced value of information functions.

A number of extensions of the current version of \pkg{hesim} would be helpful. First, we plan to extend the individual-level simulation so that users can update input data during the simulation as a function of time elapsed and/or the health states transitioned to. This would allow potentially time-varying variables such as age to update each time a patient transitions to a new state. Second, we would like to add more examples to ensure that \pkg{hesim} is fit for purpose for a wide range of applications in health economics. Third, more complex models of utility \citep{kharroubi2007modelling, hernandez2013relationship} and costs \citep{nixon2005methods} could help better capture heterogeneity in preferences and spending. Fourth, it would be useful to add functionality that automatically creates tunnel states for users in cDTSTMs. Fifth, equity considerations could be incorporated by adding support for distributional CEA \citep{asaria2015distributional, asaria2016distributional, cookson2017using}. Sixth, it would be convenient if \pkg{hesim} simulation output could be converted into formats required by other packages that can analyze the outputs of health economics models, such as those with more advanced value of information analysis capabilities. Finally, a larger aspiration is to provide support for parameterization via NMA, particularly for estimation of relative treatment effects in survival (e.g., for PSMs) and multi-state models. However, since existing \proglang{R} packages are lacking in this area, this could involve creation of a separate NMA package and a new dependency for \pkg{hesim}.

>>>>

There are a number of multi-state modeling \proglang{R} packages that can both fit models and make predictions. \pkg{hesim} leverages some of these for parameter estimation: \pkg{flexsurv} is used to fit parametric and flexible parametric survival and multi-state models when exact event times are available while \pkg{msm} is used to fit multi-state models when only panel data is available. A key difference is that \pkg{hesim} is designed specifically for health economic modeling whereas the other packages are more general purpose statistical packages.

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models \citep{de2011mstate}. It can predict state probabilities and simulate multi-state trajectories (to facilitate prediction for clock-reset models) from cumulative hazard functions evaluated at specific times. \citet{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health economic modeling since it was not designed for it. \code{mssample()}---the core function for simulating multi-state models---does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is also less computationally efficient than \pkg{hesim} for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions when efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance (see ``Benchmarks'' on the package website), we ran 1,000 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. At the time of writing, the model completed in 2.8 seconds in \pkg{hesim} and 3.1 hours in \pkg{mstate}, showing that computational considerations are not just an academic concern.

\pkg{heemod} is the only other general purpose \proglang{R} package for cost-effectiveness modeling. As noted in the introduction, there are some important differences between \pkg{hesim} and \pkg{heemod} related to the supported model types and integration of statistical and economic modeling. In addition, as with \pkg{mstate}, simulations are not vectorized, which could result in significantly reduced performance. To test this, we simulated the time-inhomogeneous total hip replacement example from \citet{briggs2006decision}. A cohort model in \pkg{heemod} completed in approximately 1.9 minutes while the same cohort model took approximately 1 second to run in \pkg{hesim} and an equivalent individual-level simulation in \pkg{hesim} ran in 7.4 seconds.

There are of course features in other \proglang{R} packages that are not in the current version of \pkg{hesim}. For example, \pkg{heemod} has an explicit function, \code{calibrate_model()}, for calibrating model parameters that are difficult to measure. Furthermore, tunnel states can be automatically created (rather than needing to be created manually) with \pkg{heemod} using the \code{state_time} variable so that costs or transition probabilities can very as a function of time in a health state. \pkg{hesim} is also limited in its support for value of information analyses. In particular, neither the expected value of partial perfect information (EVPPI) or the expected value of sample information (EVSI) can be computed with \pkg{hesim}, whereas the EVPPI can be computed with \pkg{BCEA}, \pkg{dampack}, and \pkg{SAVI}, and EVSI can be computed with \pkg{EVSI} and \pkg{dampack}. In this respect, these packages are complementary and simulation output from \pkg{hesim} could be used as inputs to their more advanced value of information functions.

A number of extensions of the current version of \pkg{hesim} would be helpful. First, we plan to extend the individual-level simulation so that users can update input data during the simulation as a function of time elapsed and/or the health states transitioned to. This would allow potentially time-varying variables such as age to update each time a patient transitions to a new state. Second, we would like to add more examples to ensure that \pkg{hesim} is fit for purpose for a wide range of applications in health economics. Third, more complex models of utility \citep{kharroubi2007modelling, hernandez2013relationship} and costs \citep{nixon2005methods} could help better capture heterogeneity in preferences and spending. Fourth, it would be useful to add options so that tunnel states could be automatically created in cDTSTMs. Fifth, equity considerations could be incorporated by adding support for distributional CEA \citep{asaria2015distributional, asaria2016distributional, cookson2017using}. Sixth, it would be convenient if \pkg{hesim} simulation output could be converted into formats required by other packages that can analyze the outputs of health economics models, such as those with more advanced value of information analysis capabilities. Finally, a larger aspiration is to provide support for parameterization via NMA, particularly for estimation of relative treatment effects in survival (e.g., for PSMs) and multi-state models. However, since existing \proglang{R} packages are lacking in this area, this could involve creation of a separate NMA package and a new dependency for \pkg{hesim}.
